\documentclass[11pt]{ucthesis}
\setlength{\oddsidemargin}{0.15in}
\def\dsp{\def\baselinestretch{1.75}\large\normalsize}
\ssp
 
\usepackage{amsmath}
\usepackage[pdftex]{graphicx}
\usepackage{times}
\usepackage{amsfonts}
\usepackage[usenames,dvipsnames]{color}
\usepackage{verbatim}
\usepackage{fancyvrb, relsize}
\usepackage{url}
\usepackage{hyperref}
\usepackage{wrapfig}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{algorithm2e}
\usepackage{listings}
\usepackage[small]{caption}


\hypersetup{%
  colorlinks=true,% hyperlinks will be black
  urlcolor=blue,
  linkcolor=blue,
  citecolor=blue,
  pdfborderstyle={/S/U/W 1}% border style will be underline of width 1pt
}

\lstdefinelanguage{pret}
{morekeywords={get_time, delay_until, delay_and_set, exception_on_expire, deactivate_exception}}


\lstset{frame=tb, 
		numberstyle=\tiny, 
		numbers=left, 
		basicstyle=\scriptsize\sffamily, 
		keywordstyle=\textbf, 
		language=pret, 
		numbersep=5pt,
		xleftmargin=8pt}
		
\captionsetup[lstlisting]{singlelinecheck=false}


\newcommand{\gettime}{\textit{get\_time}}
\newcommand{\delayuntil}{\textit{delay\_until}}
\newcommand{\delayandset}{\textit{delay\_and\_set}}
\newcommand{\exceptiononexpire}{\textit{exception\_on\_expire}}
\newcommand{\deactivateexception}{\textit{deactivate\_exception}}
\newcommand{\mtfd}{\textit{mtfd}}
\newcommand{\timerexpired}{\textit{timer\_expired}}

\newcommand{\Gettime}{\textit{Get\_time}}
\newcommand{\Delayuntil}{\textit{Delay\_until}}
\newcommand{\Delayandset}{\textit{Delay\_and\_set}}
\newcommand{\Exceptiononexpire}{\textit{Exception\_on\_expire}}
\newcommand{\Deactivateexception}{\textit{Deactivate\_exception}}

\newcommand{\platformclock}{\emph{platform clock}}

\newcommand{\rt}{real-time}
\newcommand{\Rt}{Real-time}
\newcommand{\thdint}{thread-interleaved}
\newcommand{\Thdint}{Thread-interleaved}
\newcommand{\todo}[1]{{{\textcolor{Red}{(Todo: #1)}}}}%
\newcommand{\mycaption}[1]{\vspace{-7mm}
\caption{#1}
\vspace{-4mm}}
\newcommand{\hpstar}[2]{#1}

\newcommand{\IL}[1]{{{\textcolor{Blue}{(IL: #1)}}}}
\newcommand{\HP}[1]{{{\textcolor{Purple}{(HP: #1)}}}}
\newcommand{\JR}[1]{{{\textcolor{Green}{(JR: #1)}}}}
\newcommand{\EAL}[1]{{{\textcolor{Red}{(EAL: #1)}}}}

%\renewcommand*\thesubsection{\thesection.\roman{subsection}}
\renewcommand*\thesubsubsection{\thesubsection.\alph{subsubsection}}

\begin{document}
\title{Precision Timed Machines
\iffalse
From Ptides to PtidyOS, Designing Distributed Real-Time Embedded Systems
  \titlenote{This work was supported in part by the Center for Hybrid and Embedded Software Systems (CHESS) at UC Berkeley, which receives support from the National Science Foundation (NSF awards \#0720882 (CSR-EHS: PRET), \#0931843 (ActionWebs), and \#1035672 (CSR-)CPS Ptides)), the U. S. Army Research Office (ARO \#W911NF-07-2-0019), the U. S. Air Force Office of Scientific Research (MURI \#FA9550-06-0312), the Air Force Research Lab (AFRL), the Multiscale Systems Center (MuSyC), one of six research centers funded under the Focus Center Research Program, a Semiconductor Research Corporation program, and the following companies: Bosch, National Instruments, Thales, and Toyota.}
\fi
}
\author{Isaac Suyu Liu} 
\degreeyear{2012}
\degreesemester{Spring}
\degree{Doctor of Philosophy}
\chair{Professor Edward A. Lee}
\othermembers{Professor John Wawrzynek \\
Professor Alice Agogino}
\numberofmembers{3}
%\prevdegrees{Bachelor of Science (University of California, Santa Barbara) 2007}
\field{Electrical Engineering and Computer Sciences}
\campus{Berkeley}

\maketitle
%FIXME for the final version, comment out the approval page.
%\approvalpage
\copyrightpage

\begin{abstract}
\input{abstract}
\end{abstract}

\begin{frontmatter}

\begin{dedication}
\null\vfil
{\large
\begin{center}
%\\\vspace{12pt}
To my wife, my parents, and everyone else whom I've had the privilege of running into for the first twenty-seven years of my life.
%\\\vspace{12pt}
\end{center}}
\vfil\null
\end{dedication}


\begin{acknowledgements}

I want to thank my wife, for supporting me throughout my Ph.D career, and giving me the strength to complete this work.  

\bigskip 

I want to thank my parents, for raising me and teaching me the values of perseverance and hard work.  

\bigskip

I want to thank my advisor, Edward A. Lee, for always guiding and encouraging me.    

\bigskip

I want to thank all the collaborators that worked on the PRET project with me: Ben Lickly, Hiren Patel, Jan Rieneke, Sungjun Kim, Stephen Edwards, David Broman and Mike Zimmer. This work would not have been possible without you guys.

\bigskip

I want to thank the whole ptolemy group, especially Christopher and Mary, for always taking care of everything so I could focus on my research.

\bigskip

I want to thank Jia, for providing me the template and valuable experience for the thesis writing process.

\bigskip 

I want to thank the committee members, for giving me wonderful feedback on my thesis.  

\bigskip

Finally, I want to thank everyone else not mentioned here that helped make this thesis possible.

\end{acknowledgements}

\tableofcontents
\listoffigures
\listoftables

\end{frontmatter}

\chapter{Introduction}
\label{chapter:intro}
\input{chapters/intro}

\chapter{Precision Timed Machine}
\label{chapter:pret}
\input{chapters/pret}

\chapter{Precision Timed ARM}
\label{chapter:ptarm}
\input{chapters/ptarm}

\chapter{Applications}
\label{chapter:app}
\input{chapters/1dcfd}
\input{chapters/RTencrypt}

\chapter{Related Work}
\label{chapter:related}
\input{chapters/related}

\chapter{Conclusion and Future work}
\label{chapter:summary}
\input{chapters/summary}

%backup the six stage details 
%\appendix
%\chapter{PTARM 6 stage}
%\input{chapters/ptarm6.tex}

\bibliographystyle{abbrv}
\bibliography{thesis}  

% That's all folks!
\end{document}
